Scaling up the hyperparameter grid for Naomi

### 2023-03-31 src/dev_scale-grid/scale-grid.Rmd
Use mvQuad to generate PCA grids #22

```{r}
compile("naomi_simple.cpp")
dyn.load(dynlib("naomi_simple"))

@@

```{r sd, fig.cap="Marginal standard deviations."}
sd <- sqrt(diag(sd_out$cov.fixed))

@@

```{r cov, fig.cap="Covariance matrix."}
C <- sd_out$cov.fixed
C_df <- reshape2::melt(as.matrix(C))

@@

```{r results="hide", message=FALSE}
area_merged <- read_sf(system.file("extdata/demo_areas.geojson", package = "naomi"))
pop_agesex <- read_csv(system.file("extdata/demo_population_agesex.csv", package = "naomi"))
survey_hiv_indicators <- read_csv(system.file("extdata/demo_survey_hiv_indicators.csv", package = "naomi"))
art_number <- read_csv(system.file("extdata/demo_art_number.csv", package = "naomi"))
anc_testing <- read_csv(system.file("extdata/demo_anc_testing.csv", package = "naomi"))
pjnz <- system.file("extdata/demo_mwi2019.PJNZ", package = "naomi")
spec <- naomi::extract_pjnz_naomi(pjnz)

@@

Some hyperparameters have higher standard deviation than others in part due to the different scales that they are on.
For example if the hyperparameter is on the logit scale then it looks as though it will have a higher standard deviation than another those on the log scale (Figure \@ref(fig:sd)).
The average standard deviation for hyperparameters starting with `"log_"` is `r data.frame(par = names(sd), sd = unname(sd)) %>% filter(str_detect(par, "^log_")) %>% summarise(mean_sd = mean(sd)) %>% pull(mean_sd)` as compared with `r data.frame(par = names(sd), sd = unname(sd)) %>% filter(str_detect(par, "^logit_")) %>% summarise(mean_sd = mean(sd)) %>% pull(mean_sd)` for those starting with `"logit_"`.

@@

The relative contributions of each principle component are:

```{r scree, fig.cap="Scree plot."}
tv_df <- data.frame(
  n = 1:length(lambda),
  tv = cumsum(lambda / sum(lambda))
)

@@

```{r pc-loadings, fig.cap="PC loadings."}
reshape2::melt(as.matrix(E)) %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  labs(x = "", y = "", fill = "E[i, j]") +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "PC loading", y = "Hyper") +
  theme_minimal()
```

@@

```{r cov-reconstruction, fig.cap="PC loadings."}
E_s <- E[, 1:s]
Lambda_s <- Lambda[1:s, 1:s]
C_s <- E_s %*% Lambda_s %*% t(E_s)

@@

Let $a \in \mathbb{R}^d$ and $b \in \mathbb{R}^s$.
The PCA provides a mapping $f: \mathbb{R}^d \to \mathbb{R}^s$ which compresses each $a$ to a smaller $b$ given by $f(a) = (E_s \Lambda_s^{1/2})^\top a = b$.
Then $f^{-1}$ is given by $f^{-1}(b) = E_s \Lambda_s^{-1/2} b = a$, using the fact that $E_s^\top = E_s^{-1}$.
So we need to multiply our nodes by $E_s \Lambda_s^{- 1/2}$ as follows:

```{r}
a_nodes <- mvQuad::getNodes(gg_s)
b_nodes <- t(E_s %*% diag(lambda[1:s]^{-0.5}) %*% t(a_nodes))

dim(a_nodes)
dim(b_nodes)
```

Put this into a function, which also adjusts according to the mean and reweights the nodes:

```{r}
#' @param m Mean vector
#' @param C Covariance matrix
#' @param s Small grid dimension
#' @param k Number of points per small grid dimension
pca_rescale <- function(m, C, s, k) {
  d <- nrow(C)
  stopifnot(d == length(m))
  eigenC <- eigen(C)
  lambda <- eigenC$values
  Lambda <- diag(lambda)
  E <- eigenC$vectors
  E_s <- E[, 1:s] 
  gg_s <- mvQuad::createNIGrid(dim = s, type = "GHe", level = k) 
  nodes_out <- t(E_s %*% diag(lambda[1:s]^{-0.5}, ncol = s) %*% t(mvQuad::getNodes(gg_s)))
  for(j in 1:d) nodes_out[, j] <- nodes_out[, j] + m[j]
  weights_out <- mvQuad::getWeights(gg_s) * as.numeric(mvQuad::getWeights(mvQuad::createNIGrid(dim = d - s, type = "GHe", level = 1)))

@@

```{r gaussian-2d, class.source = 'fold-hide', fig.cap="Gaussian 2D objective function."}
mu <- c(1, 1.5)
cov <- matrix(c(2, 1, 1, 1), ncol = 2)

obj <- function(theta) {
  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov, log = TRUE)
}

@@

```{r}
pca_grid_1 <- pca_rescale(m = mu, C = cov, s = 1, k = 1)
pca_grid_3 <- pca_rescale(m = mu, C = cov, s = 1, k = 3)
pca_grid_5 <- pca_rescale(m = mu, C = cov, s = 1, k = 5)
pca_grid_7 <- pca_rescale(m = mu, C = cov, s = 1, k = 7)
```

For example, the version with 7 points looks as follows:

```{r pca-grid-7, fig.cap="AGHQ-PCA grid for Gaussian with $k = 7$."}
plot_points <- function(gg) {
  plot +
    geom_point(
      data = mvQuad::getNodes(gg) %>%
        as.data.frame() %>%
        mutate(weights = mvQuad::getWeights(gg)),
      aes(x = V1, y = V2, size = weights),
      alpha = 0.8,
      col = multi.utils::cbpalette()[2],
      inherit.aes = FALSE
    ) +
    scale_size_continuous(range = c(1, 2)) +
    labs(x = "", y = "", size = "Weight") +
    theme_minimal()
}

plot_points(pca_grid_7)
```

@@

```{r}
opt_bfgs <- aghq::optimize_theta(ff, mu, control = default_control(method = "BFGS"))
opt_bfgs$mode
opt_bfgs$hessian
```

@@

@@

```{r fig.cap="Unit test for the Gaussian example"}
truelognormconst <- 0

results <- data.frame(
  "method" = c("Truth", paste0("aghq, k = ", c(1, 3, 5, 7)), paste0("aghq-pca, k = ", c(1, 3, 5, 7))),
  "lognormconst" = c(truelognormconst, sapply(norm_bfgs, function(x) x$lognormconst), sapply(norm_pca_bfgs, function(x) x$lognormconst)),
  "type" = as.factor(c("Truth", rep("AGHQ", 4), rep("AGHQ-PCA", 4)))
)

@@

```{r}
grid_bfgs_1 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 1)
mvQuad::rescale(grid_bfgs_1, m = m, C = C)

@@

pca_grid_bfgs_1 <- pca_rescale(m = m, C = C, s = 1, k = 1)
pca_grid_bfgs_3 <- pca_rescale(m = m, C = C, s = 1, k = 3)
pca_grid_bfgs_5 <- pca_rescale(m = m, C = C, s = 1, k = 5)
pca_grid_bfgs_7 <- pca_rescale(m = m, C = C, s = 1, k = 7)

#' AGHQ
plot_points(grid_bfgs_7)

#' PCA-AGHQ
#' Something to fix here: points not being scaled to the right place
plot_points(pca_grid_bfgs_7)

@@

results <- data.frame(
  "method" = c("Truth", paste0("aghq, k = ", c(1, 3, 5, 7)), paste0("aghq-pca, k = ", c(1, 3, 5, 7))),
  "lognormconst" = c(truelognormconst, sapply(norm_bfgs, function(x) x$lognormconst), sapply(norm_pca_bfgs, function(x) x$lognormconst)),
  "type" = as.factor(c("Truth", rep("AGHQ", 4), rep("AGHQ-PCA", 4)))
)

>>>>

```{r, class.source = 'fold-hide'}
compile("naomi_simple.cpp")
dyn.load(dynlib("naomi_simple"))

@@

```{r sd, class.source = 'fold-hide', fig.cap="Marginal standard deviations."}
sd <- sqrt(diag(sd_out$cov.fixed))

@@

```{r cov, class.source = 'fold-hide', fig.cap="Covariance matrix."}
C <- sd_out$cov.fixed
C_df <- reshape2::melt(as.matrix(C))

@@

```{r, class.source = 'fold-hide', results="hide", message=FALSE}
area_merged <- read_sf(system.file("extdata/demo_areas.geojson", package = "naomi"))
pop_agesex <- read_csv(system.file("extdata/demo_population_agesex.csv", package = "naomi"))
survey_hiv_indicators <- read_csv(system.file("extdata/demo_survey_hiv_indicators.csv", package = "naomi"))
art_number <- read_csv(system.file("extdata/demo_art_number.csv", package = "naomi"))
anc_testing <- read_csv(system.file("extdata/demo_anc_testing.csv", package = "naomi"))
pjnz <- system.file("extdata/demo_mwi2019.PJNZ", package = "naomi")
spec <- naomi::extract_pjnz_naomi(pjnz)

@@

Some hyperparameters have higher standard deviation than others in part due to the different scales that they are on.
For example if the hyperparameter is on the logit scale then it looks as though it will have a higher standard deviation than another those on the log scale (Figure \@ref(fig:sd)).
The average standard deviation for hyperparameters starting with `"log_"` is `r round(data.frame(par = names(sd), sd = unname(sd)) %>% filter(str_detect(par, "^log_")) %>% summarise(mean_sd = mean(sd)) %>% pull(mean_sd), 3)` as compared with `r round(data.frame(par = names(sd), sd = unname(sd)) %>% filter(str_detect(par, "^logit_")) %>% summarise(mean_sd = mean(sd)) %>% pull(mean_sd), 3)` for those starting with `"logit_"`.

@@

The relative contributions of each principle component are given by $\lambda_i / \sum_i \lambda_i$:

```{r scree, class.source = 'fold-hide', fig.cap="Scree plot."}
tv_df <- data.frame(
  n = 1:length(lambda),
  tv = cumsum(lambda / sum(lambda))
)

@@

```{r pc-loadings, class.source = 'fold-hide', fig.cap="PC loadings."}
reshape2::melt(as.matrix(E)) %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  labs(x = "", y = "", fill = "E[i, j]") +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "PC loading", y = "Hyper") +
  theme_minimal()
```

@@

```{r cov-reconstruction, class.source = 'fold-hide', fig.cap="PC loadings."}
E_s <- E[, 1:s]
Lambda_s <- Lambda[1:s, 1:s]
C_s <- E_s %*% Lambda_s %*% t(E_s)

@@

Create a function to do the PCA rescaling, which also adapts according to the mean and reweights the nodes:

```{r}
#' @param m Mean vector
#' @param C Covariance matrix
#' @param s Small grid dimension
#' @param k Number of points per small grid dimension
pca_rescale <- function(m, C, s, k) {
  d <- nrow(C)
  stopifnot(d == length(m))
  eigenC <- eigen(C)
  lambda <- eigenC$values
  Lambda <- diag(lambda)
  E <- eigenC$vectors
  E_s <- E[, 1:s] 
  gg_s <- mvQuad::createNIGrid(dim = s, type = "GHe", level = k) 
  nodes_out <- t(E_s %*% diag(lambda[1:s]^{0.5}, ncol = s) %*% t(mvQuad::getNodes(gg_s)))
  for(j in 1:d) nodes_out[, j] <- nodes_out[, j] + m[j]
  weights_out <- mvQuad::getWeights(gg_s) * as.numeric(mvQuad::getWeights(mvQuad::createNIGrid(dim = d - s, type = "GHe", level = 1)))

@@

```{r gaussian-2d, class.source = 'fold-hide', fig.cap="Gaussian 2D objective function."}
m <- c(1, 1.5)
C <- matrix(c(2, 1, 1, 1), ncol = 2)

obj <- function(theta) {
  mvtnorm::dmvnorm(theta, mean = m, sigma = C, log = TRUE)
}

@@

```{r}
pca_grid_1 <- pca_rescale(m = m, C = C, s = 1, k = 1)
pca_grid_3 <- pca_rescale(m = m, C = C, s = 1, k = 3)
pca_grid_5 <- pca_rescale(m = m, C = C, s = 1, k = 5)
pca_grid_7 <- pca_rescale(m = m, C = C, s = 1, k = 7)
```

For example, the version with 3 points looks as follows:

```{r pca-grid-3, class.source = 'fold-hide', fig.cap="AGHQ-PCA grid for Gaussian with $k = 3$."}
plot_points <- function(gg) {
  plot +
    geom_point(
      data = mvQuad::getNodes(gg) %>%
        as.data.frame() %>%
        mutate(weights = mvQuad::getWeights(gg)),
      aes(x = V1, y = V2, size = weights),
      alpha = 0.8,
      col = multi.utils::cbpalette()[2],
      inherit.aes = FALSE
    ) +
    scale_size_continuous(range = c(1, 2)) +
    labs(x = "", y = "", size = "Weight") +
    theme_minimal()
}

plot_points(pca_grid_3)
```

@@

```{r}
opt_bfgs <- aghq::optimize_theta(ff, m, control = default_control(method = "BFGS"))
opt_bfgs$mode
opt_bfgs$hessian
Matrix::forceSymmetric(solve(opt_bfgs$hessian))
```

@@

And a second way of producing the AGHQ-PCA grids directly using `mvQuad`:

```{r, fig.cap="Alternative approach AGHQ-PCA grid for Gaussian with $k = 3$."}
pca2_grid_1 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(1, 1))
mvQuad::rescale(pca2_grid_1, m = m, C = C, dec.type = 1)

pca2_grid_3 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(3, 1))
mvQuad::rescale(pca2_grid_3, m = m, C = C, dec.type = 1)

pca2_grid_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(5, 1))
mvQuad::rescale(pca2_grid_5, m = m, C = C, dec.type = 1)

pca2_grid_7 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(7, 1))
mvQuad::rescale(pca2_grid_7, m = m, C = C, dec.type = 1)

plot_points(pca2_grid_3)

norm_pca2_bfgs_1 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_1, adapt = FALSE)
norm_pca2_bfgs_3 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_3, adapt = FALSE)
norm_pca2_bfgs_5 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_5, adapt = FALSE)
norm_pca2_bfgs_7 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_7, adapt = FALSE)
norm_pca2_bfgs <- list(norm_pca2_bfgs_1, norm_pca2_bfgs_3, norm_pca2_bfgs_5, norm_pca2_bfgs_7)
```

@@

```{r, class.source = 'fold-hide', fig.cap="Unit test for the Gaussian example"}
truelognormconst <- 0

results <- data.frame(
  "method" = c("Truth", paste0("aghq, k = ", c(1, 3, 5, 7)), paste0("aghq-pca, k = ", c(1, 3, 5, 7)), paste0("aghq-pca2, k = ", c(1, 3, 5, 7))),
  "lognormconst" = c(truelognormconst, sapply(c(norm_bfgs, norm_pca_bfgs, norm_pca2_bfgs), function(x) x$lognormconst)),
  "type" = c("Truth", rep("AGHQ", 4), rep("AGHQ-PCA", 4), rep("AGHQ-PCA2", 4))
)

ggplot(results, aes(x = method, y = lognormconst, col = as.factor(type))) +
  geom_point(size = 1) +
  labs(x = "Method", y = "Log normalising constant", col = "") +
  scale_color_manual(values = multi.utils::cbpalette()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70, vjust = 0.5, hjust = 0.5))
```

@@

```{r, fig.cap="AGHQ grid for non-Gaussian with $k = 3$."}
grid_bfgs_1 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 1)
mvQuad::rescale(grid_bfgs_1, m = m, C = C)

@@

plot_points(grid_bfgs_3)
```

```{r, fig.cap="AGHQ-PCA grid for non-Gaussian with $k = 3$."}
pca_grid_bfgs_1 <- pca_rescale(m = m, C = C, s = 1, k = 1)
pca_grid_bfgs_3 <- pca_rescale(m = m, C = C, s = 1, k = 3)
pca_grid_bfgs_5 <- pca_rescale(m = m, C = C, s = 1, k = 5)
pca_grid_bfgs_7 <- pca_rescale(m = m, C = C, s = 1, k = 7)

@@

plot_points(pca_grid_bfgs_3)
```

```{r, fig.cap="Alternative approach AGHQ-PCA grid for non-Gaussian with $k = 3$."}
pca2_grid_1 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(1, 1))
mvQuad::rescale(pca2_grid_1, m = m, C = C, dec.type = 1)

pca2_grid_3 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(3, 1))
mvQuad::rescale(pca2_grid_3, m = m, C = C, dec.type = 1)

pca2_grid_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(5, 1))
mvQuad::rescale(pca2_grid_5, m = m, C = C, dec.type = 1)

pca2_grid_7 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(7, 1))
mvQuad::rescale(pca2_grid_7, m = m, C = C, dec.type = 1)

norm_pca2_bfgs_1 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_1, adapt = FALSE)
norm_pca2_bfgs_3 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_3, adapt = FALSE)
norm_pca2_bfgs_5 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_5, adapt = FALSE)
norm_pca2_bfgs_7 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_7, adapt = FALSE)
norm_pca2_bfgs <- list(norm_pca2_bfgs_1, norm_pca2_bfgs_3, norm_pca2_bfgs_5, norm_pca2_bfgs_7)

plot_points(pca2_grid_3)
```

```{r, class.source = 'fold-hide', fig.cap="Unit test for the non-Gaussian example"}
results <- data.frame(
  "method" = c("Truth", paste0("aghq, k = ", c(1, 3, 5, 7)), paste0("aghq-pca, k = ", c(1, 3, 5, 7)), paste0("aghq-pca2, k = ", c(1, 3, 5, 7))),
  "lognormconst" = c(truelognormconst, sapply(c(norm_bfgs, norm_pca_bfgs, norm_pca2_bfgs), function(x) x$lognormconst)),
  "type" = as.factor(c("Truth", rep("AGHQ", 4), rep("AGHQ-PCA", 4), rep("AGHQ-PCA2", 4)))
)
